Physcraper: a python package for continual update of evolutionary estimates using the Open Tree of Life

### 2020-08-28 methods-extended.Rmd
extended methods

---
title: 'Physcraper: Body'
author:
  - Luna L. Sanchez Reyes
  - Martha Kandziora
  - Emily Jane McTavish
  - University of California, Merced
date: '`r format(Sys.time(), "%d %B, %Y")`'
bibliography: paper.bib
csl: apa.csl
output:
  redoc::redoc
---

# Introduction

Phylogenetic estimates of evolutionary relationships capture the shared history of living organisms and provide key context for all our biological observations.
Public biological databases constitute an amazing resource for evolutionary estimation, but a large portion of molecular data publicly available has never been incorporated into any phylogenetic estimate. Extending existing phylogenetic estimates with new DNA sequence data, geographical location, and other metadata in a reproducible and continuous manner is possible by automating connections between biological databases. Here, we introduce Physcraper, a tool that uses existing phylogenetic research from public biological databases to update a starting tree and single locus alignments. Physcraper lets researchers build upon molecular data that taxon specialists have assessed and deemed appropriate for a specific phylogenetic scope.

The prevalence of taxonomic idiosyncrasies across databases represent a key challenge to automatically connecting data from disparate biological databases in a phylogenetic context. To standardize taxonomic names, a unified system is needed. The main aim of the Open Tree of Life project (OpenTree)
<!-- [project](https://opentreeoflife.github.io/)  -->
is to construct a comprehensive, dynamic and digitally-available tree of life by synthesizing published phylogenetic trees along with taxonomic data. Currently, OpenTree's "synthetic" tree comprises 2.3 million tips, of which around 90,000 are represented by phylogenetic estimates - the remaining 1.4 million taxa are placed in the tree based on their taxonomic assignment. To achieve this, OpenTree unifies taxonomy from various databases [@rees2017automated] such as the USA National Center for Biodiversity Information (NCBI) molecular database GenBank [@benson2000genbank; @wheeler2000database], the Global Biodiversity Information Facility [GBIF; @secretariat2017gbif], and the World Register of Marine Species [WoRMS; www.marinespecies.org/]. This provideslinks taxonomic information across databases a key resource that can be used to connect data from virtually any biological database to phylogenetic data that has been standardized to OpenTree's unified taxonomy.

Another challenge to incorporating molecular data from public databases to update phylogenetic knowledge is assembling high-quality homology hypotheses.
Species tree reconstructions from multiple single locus data sets taking into account the multispecies coalescent model are seen as the gold standard for inferring species relationships [@song2012resolving].
Genomics has, and will continue to, revolutionize phylogenetic inference.
Yet, different research questions call for different genomic sequencing approaches, from whole genomes, to transcriptomes, restriction-site associated DNA sequencing, single nucleotide polymorphisms, microsatellites, and ultra-conserved elements, which has lead to largely non-overlapping genomic data sets across taxa, creating difficulty in wide scale phylogenetic reconstructions.
<!-- Also, as @andermann2020guide point out, "assembling full genomes is often unnecessary for phylogenomic studies if the main goal is to retrieve an appropriate number of phylogenetically informative characters from several independent and single-copy genetic markers [@jones2016targeted]." -->
While phylogenomics ameliorates the problem of non-overlapping genomic data sets by focusing on targeted capture of informative characters from independent and single-copy genetic markers [@andermann2020guide; @jones2016targeted], decades of single locus sequencing have already generated homologous DNA data sets that can be used for phylogenetic reconstruction at many scales.

Indeed, more than a decade ago, GenBank release number 159 (April 15, 2007) already hosted 72 million DNA sequences. These sequences were gauged to have the potential to resolve phylogenetic relationships of most (`r round(236023/240708, digits=4)*100`%) of the almost 241, 000 distinct taxa in the NCBI taxonomy at the time [@sanderson2008phylota]. Assembling a DNA alignment from such a massive database can be done "by hand", but it requires huge amounts of time and it is mostly a non-reproducible approach. Computational pipelines that make DNA sequence search faster and more efficient, as well as more reproducible, have been applied to study evolutionary relationships among a variety of organisms [e.g., @smith2009mega; @antonelli2017toward; @izquierdo2014pumper].
<!-- *maybe a box of all pipelines?* -->
However, even in phylogenomic reconstructions, thoughtfully curated markers and alignments can improve phylogenetic reconstructions [@fragoso2017pilot].

A way to incorporate the best of two worlds (massive amounts of newly released molecular data and fine-grained curation from human experts) is to rely on published manually curated homology hypotheses as "jump-start" alignments [@morrison2006multiple]. The TreeBASE database [@piel2009treebase] hosts about 8, 200 publicly accessible alignments, providing information on evolutionary relationships of around 100, 000 distinct taxa [(see TreeBASE's website about)](https://www.treebase.org/treebase-web/home.html#:~:text=TreeBASE%20is%20produced%20and%20governed,mapped%20to%20104%2C593%20distinct%20taxa.), representing a source of valuable expert knowledge. Linking published alignments with public molecular data that has not yet been included in any public phylogenetic estimate, has the potential to accelerate the enrichment and updating of phylogenetic relationships in many regions of the tree of life. The phylogenies associated with TreeBASE alignments have been integrated to the OpenTree's datastore, the Phylesystem [@mctavish2015phylesystem], and metadata linking them to their corresponding alignment repository is available, providing a non-automated way of linking trees with the alignment that generated them.

Physcraper relies on programmatic access protocols (API's) available to automatically link molecular data from GenBank to alignments from TreeBASE and phylogenies from OpenTree's Phylesystem, to continually update and enrich phylogenetic knowledge based on expertly-curated homology hypotheses. Physcraper also provides new types of access to various general OpenTree programmatic tools for comparison of existing phylogenetic hypotheses with newly generated ones.
Physcraper is coded as a Python pipeline that focuses on data interoperability, by integrating taxonomic name matching across biological databases. This integration also allows users to rapidly place new data from a diverse range of biological databases in an evolutionary context, making a variety of downstream analyses straightforward.

# The Physcraper framework

```{r framework, echo=FALSE, fig.cap="The Physcraper framework consists of 4 steps (see text). The software is fully described on its documentation website at physcraper.readthedocs.io, along with installation instructions, function usage descriptions, examples and tutorials.", out.width = '85%', fig.retina= 2, fig.align = "center"}
# knitr::include_graphics("https://raw.githubusercontent.com/McTavishLab/physcraper/pyopensci/docs/img/schematic.svg")
knitr::include_graphics("docs/figs/schematic.pdf")
```

The general Physcraper framework is depicted in Figure \@ref(fig:framework). Briefly, it consists of 4 steps: 1) identifying and processing a phylogenetic tree to update and its underlying alignment; 2) performing a constrained BLAST search of sequences from the original alignment on the GenBank DNA database, and filtering of newly found sequences; 3) profile-aligning new sequences that passed the filtering to the original alignment; 4) performing a phylogenetic analysis and comparing the updated tree to previous phylogenetic estimates in the focus group. Next, we will describe technical details for each step.

@@

In order to take advantage of the OpenTree tools, it is reccommended that the input tree is either stored in the
  OpenTree [Phylesystem](https://github.com/opentreeoflife/phylesystem), or submitted via OpenTree's curator [application](https://tree.opentreeoflife.org/curator) [@mctavish2015phylesystem].
  Currently, only trees connected to a published study can be stored in the Phylesystem.
  Users can choose from among the 2, 950 studies in OpenTree's Phylesystem that have alignments on TreeBASE.
  If the user is not ready to make the input tree public, tree tip labels can be standardized to the unified OpenTree taxonomy using OpenTree's bulk Taxonomic Name Resolution Service [TNRS](https://tree.opentreeoflife.org/curator/tnrs/) tool.
  This step is referred to as taxonomic name mapping and Phylesystem stored trees are processed in this way upon submission.
  Physcraper saves a summary "csv" file with results from the taxon name standardization for advantage of the user, including the mappings to unique identifiers in the OpenTree and NCBI taxonomies.
  If taxon names can't be mapped, their taxonomic information is not used in comparison analysis. These taxa will still be used in the sequence search and phylogenetic reconstruction steps.
  Mapping tip names to OpenTree's unified taxonomy saves a set of user defined characteristics
  that are essential for automatizing the phylogeny updating process. The most relevant of these is the standardized taxonomic names and the definition of ingroup and outgroup taxa, allowing to automatically set the root for the updated tree on the final steps of the pipeline.

@@

# Example: The hollies

To illustrate the utility of Physcraper we propose a scenario in which a user is interested in phylogenetic relationships within the genus *Ilex*. Commonly known as "hollies", the genus encompasses between 400-700 living species, and is the only extant clade within the family Aquifoliaceae, order Aquifoliales of flowering plants.

An online literature review in June 2020 (google scholar search for "ilex phylogeny") reveals that there are several published phylogenetic trees showing relationships within the hollies [@cuenoud2000molecular; @manen2010history; @setoguchi2000intersectional; @selbach2009new], but only two have their data available publicly [@gottlieb2005molecular; @yao2020phylogeny].
@gottlieb2005molecular made original tree and alignment data available in [TreeBASE](https://treebase.org/treebase-web/search/study/summary.html?id=1091). The "Gottlieb2005" tree sampling 41 species was added to the [OpenTree Phylesystem](https://tree.opentreeoflife.org/curator/study/edit/pg_2827/?tab=home) and its information has been integrated into [OpenTree's synthetic tree](https://devtree.opentreeoflife.org/opentree/opentree12.3@mrcaott68451ott89474/Ilex-theizans--Ilex-dumosa).

The most recent *Ilex* tree from @yao2020phylogeny, has been made available in the [OpenTree Phylesystem](https://tree.opentreeoflife.org/curator/study/view/ot_1984) and in the [DRYAD repository](https://datadryad.org/stash/dataset/doi:10.5061/dryad.k0p2ngf4x). The "Yao2020" tree, is the best sampled phylogenetic tree yet available for the hollies, with 175 tips.
This makes it a great case to highlight the functionalities of Physcraper and assess its limitations.
A tutorial as well as illustrated examples of functions implemented on each step of the analysis are available in Physcraper's documentation website.
Figure \@ref(fig:results) shows results from a Physcraper analysis applied to an alignment of internal transcribed spacer DNA regions (ITS) from @gottlieb2005molecular. Physcraper ran on a local BLAST database, on a laptop Linux computer for 19hrs 45min to perform BLAST and RAxML analyses, with bootstrap analyses taking an additional 13hrs. **MTH: you probably need some details
252 about the hardware, given the fact that you are discussing running times**
The updated Gottlieb2005 tree contains all 41 distinct taxa from the original study plus 231 new tips, contributing phylogenetic data to 84 additional taxa within *Ilex*. The best RaxML tree is 99% resolved, while the consensus tree is 75% resolved, meaning that 25% of nodes have a bootstrap < 0.1, while 48% nodes have bootstrap support > 0.75.
A large portion of internal branches are negligibly small, with 30 branches < 0.00001 substitution rate units, from which only 9 have a bootstrap support > 0.75 (Fig. \@ref(fig:results)).
For comparison, the Yao2020 tree also contains all 41 distinct taxa from the original Gottlieb2005 study,
<!-- has less new tips (134), but  -->
and contributes phylogenetic data to 134 additional taxa within *Ilex*, from which
`r 135-68` are also in the updated Gottlieb2005 tree. @yao2020phylogeny also used ITS as a marker, but their data in GenBank is not public yet, so Physcraper was unable to incorporate 68 additional taxa into the analysis.
When those sequences are available online, repeating the physcraper run will easily incorporate them into the expanded analysis.
The Yao2020 tree also lacks 18 taxa that were added by Physcraper.
<!-- **Any statements about relationships or anything? Are there multiple seqs for these taxa?**
add 1 sentence saying that it is odd that we are getting differnet relationships given what is likely to be the same data
but we can't compare the exact seqs or alignemnt, bc they are not available.
 -->
This might be caused by the method they used to download existing ITS *Ilex* sequences from GenBank, which is not fully explained in the publication.
  Other examples of Physcraper runs on empirical data at a range of taxonomic scales are available in the Physcraper documentation online.
<!-- The alignments have THIS MANY columns and THIS AMOUNT of missing data. -->
<!-- RF DISTANCE INTERPRETATION -->
<!-- ADD ML ESTIMATES OF UPDATED TREE VS ORIGINAL TREE ?-->

<!-- ```{bash echo= FALSE, include=FALSE, message= FALSE, results='hide'}
wget -O docs/figs/ilex-plot-tips-and-branches-1.png https://raw.githubusercontent.com/McTavishLab/physcraperex/master/docs/articles/ilex_files/figure-html/plot-tips-and-branches-1.png
``` -->

```{r results, echo=FALSE, fig.cap="A) Phylogenetic tree obtained by updating the Gottlieb et al. 2005 tree in B) using Physcraper.", out.width = '100%', fig.retina= 2}

# knitr::include_graphics("docs/figs/ilex-plot-tips-and-branches-1.png")
knitr::include_graphics("docs/figs/ilex-results.png")

```

\newpage

*Figure \@ref(fig:results) caption continued*: Tips in original alignment and new tips added with Physcraper are depicted in black and red, respectively. Physcraper obtained sequences from the GenBank database via local BLAST of all sequences in the original alignment that generated tree in B), filtered them following criteria specified in section "DNA sequence search and filtering", aligned them to the original alignment using MUSCLE and performed a phylogenetic reconstruction using RAxML with 100 bootstraps. B-D show results of the conflict analysis performed with OpenTree tools.

# Discussion

Data repositories designed to preserve and democratize access to biological data
constitute an essential resource for evolutionary estimation. While data keep accumulating,
our ability to reanalyze and incorporate new data with available knowledge is being challenged.

Various phylogenetic pipelines have been designed to make evolutionary sense of the vast amount of public molecular data available in different ways.
Pipelines like Phylota [@sanderson2008phylota], PyPHLAWD [@smith2009mega], and SUPERSMART [@antonelli2017toward]
focus mainly on generating full trees *de novo* (i.e., inferring phylogenetic relationships from a newly generated homology hypothesis, as opposed to e.g., supertrees, that are generated by assembling previous phylogenetic estimates) for regions of the Tree of
Life that have no phylogenetic assessment yet in published studies.
While Physcraper does not generate phylogenies *de novo* in a traditional sense,
it successfully generates new phylogenetic knowledge, revealing the potential of pre-existing
phylogenetic knowledge available in tree repositories to facilitate phylogenetic placement of public molecular data.
The PUMPER pipeline [@izquierdo2014pumper] also uses the concept of updating
pre-existing alignments to incorporate public molecular data into phylogenies. Unfortunately, installation of the tool was unsuccesful following instructions from the author, and we were unable to conduct a comparison analysis to Physcraper. **MTH: add some more details about what aspect of the PUMPER install failed**
Physcraper on it's own generates individual gene trees, which do not capture the full complexity of species' evolutionary history [@song2012resolving]. However, using Physcraper it is straightforward to gather alignments and gene trees for multiple loci for a taxonomic group of interest, which can be used as inputs for downstream tools to perform coalescent analyses and assess species tree relationships, (e.g. ASTRAL [@mirarab2014astral], BEAST2 [@bouckaert2019beast], SVD Quartets [@chifman2014quartet]).

As far as we know, Physcraper is the only phylogenetic pipeline that links molecular and tree data repositories.
It can also link to various biological databases, such as GBIF, the Paleobiology Database, and others, by leveraging the integrated taxonomy from OpenTree [@rees2017automated], a resource that is contributing to the Encyclopedia of Life main goal of "exploring and analysing biodiversity at an accelerated pace, and returning systematics into the mainstream of science" [@wilson2003encyclopedia].

As such, Physcraper can be used to rapidly (in a matter of hours)
solve challenges overarching both fields of ecology and evolution, such as
placing newly discovered species phylogenetically [@webb2010biodiversity],
obtaining trees for ecophylogenetic studies [@helmus2012phylogenetic],
systematizing molecular (and other) databases, i.e., curating taxonomic assignations [@san2010molecular],
and generating custom species trees for ecological and evolutionary downstream analyses [@stoltzfus2013phylotastic].

Data repositories hold more information than meets the eye.
Besides the main data, they are rich sources of metadata that can be leveraged for the advantage of all areas of biology as well as the advancement of scientific policy and applications.
Usually, initial ideas about the data are changed by new analyses.
Ideally, new understanding of the data can be continually registered on databases
to consistently expose newcomers to the most up to date knowledge about the data.

# Acknowledgements

Research was supported by the grant "Sustaining the Open Tree of Life", National Science Foundation ABI No. 1759838, and ABI No. 1759846.
Computer time was provided by the Multi-Environment Research Computer for Exploration and Discovery (MERCED) cluster from the University of California, Merced (UCM), supported by the NSF Grant No. ACI-1429783.

We thank the members of the OpenTree development team and the "short bar" Science and Engineering Building 1, UCM, joint lab paper discussion group for valuable comments on this manuscript.

The authors have no conflict of interest to declare.

# Authors' Contributions

LLSR wrote the manuscript, alignment code, documentation, performed analyses and developed examples; MK wrote code for ncbidataparser module, filtering of sequences per OTU and using offline blast searches, wrote documentation and tests; EJM conceived study, wrote most of the code, documentation and tests.
All authors contributed to the manuscript and gave final approval for publication.

# Data Archiving

Physcraper source code available at https://github.com/McTavishLab/physcraper

Documentation available at https://physcraper.readthedocs.io/en/latest/index.html

Illustrated examples available at https://github.com/McTavishLab/physcraperex

This is a reproducible manuscript available at https://github.com/McTavishLab/physcraper_ms

\newpage
# References

>>>>

---
output:
  html_document:
    keep_md: yes
    toc: false
geometry: margin = 1in
bibliography: paper.bib
csl: submission-mee/methods-in-ecology-and-evolution.csl
# csl: submission_sysbio/systematic-biology.csl
link-citations: yes
linestretch: 2
---

# The Physcraper Framework

The general Physcraper framework consists of 4 general steps: **1) identifying and processing** a phylogenetic **tree to update and its underlying alignment**; **2) performing a constrained BLAST search** of sequences from the original alignment on the GenBank DNA database, and filtering of newly found sequences; **3) profile-aligning new sequences** that passed the filtering to the original alignment; **4) performing a phylogenetic analysis and comparing the updated tree** to previous phylogenetic estimates in the focus group. Next, we will describe each step with some technical detail.

@@

In order to take advantage of the OpenTree tools, it is recommended that the input tree is either stored in the
  OpenTree [Phylesystem](https://github.com/opentreeoflife/phylesystem), or submitted via OpenTree's curator [application](https://tree.opentreeoflife.org/curator) [@mctavish2015phylesystem].
  Currently, only trees connected to a published study can be stored in the Phylesystem.
  Users can choose from among the 2, 950 studies in OpenTree's Phylesystem that have alignments on TreeBASE.
  If the user is not ready to make the input tree public, tree tip labels can be standardized to the unified OpenTree taxonomy using OpenTree's bulk Taxonomic Name Resolution Service [TNRS](https://tree.opentreeoflife.org/curator/tnrs/) tool.
  This step is referred to as taxonomic name mapping and Phylesystem stored trees are processed in this way upon submission.
  Physcraper saves a summary "csv" file with results from the taxon name standardization for advantage of the user, including the mappings to unique identifiers in the OpenTree and NCBI taxonomies.
  If taxon names can't be mapped, their taxonomic information is not used in comparison analysis. These taxa will still be used in the sequence search and phylogenetic reconstruction steps.
  Mapping tip names to OpenTree's unified taxonomy saves a set of user defined characteristics
  that are essential for automatizing the phylogeny updating process. The most relevant of these is the standardized taxonomic names and the definition of ingroup and outgroup taxa, allowing to automatically set the root for the updated tree on the final steps of the pipeline.

@@

# References
